---
title: "Week 6 Assignment"
author: "Jordan Erickson"
date: "Monday, October 06, 2014"
output: html_document
---


Before installing `rvest`, I need to install `devtools` to get the `rvest` package from github.

```{r install_devtools}
# install.packages("devtools")

library(devtools)
# build_github_devtools()

#### Restart R before continuing ####
# install.packages("devtools.zip", repos = NULL)

# Remove the package after installation
# unlink("devtools.zip")
```


Install `rvest` package from github.

```{r install_rvest}
install_github("hadley/rvest")
library(rvest)
```


Run sample code to verify `rvest` loaded properly.

```{r verify}
lego_movie <- html("http://www.imdb.com/title/tt1490017/")

rating <- lego_movie %>% 
  html_nodes("strong span") %>%
  html_text() %>%
  as.numeric()
rating
#> [1] 7.9

cast <- lego_movie %>%
  html_nodes("#titleCast .itemprop span") %>%
  html_text()
cast
#>  [1] "Will Arnett"     "Elizabeth Banks" "Craig Berry"     "Alison Brie"    
#>  [5] "David Burrows"   "Anthony Daniels" "Charlie Day"     "Amanda Farinos" 
#>  [9] "Keith Ferguson"  "Will Ferrell"    "Will Forte"      "Dave Franco"    
#> [13] "Morgan Freeman"  "Todd Hansen"     "Jonah Hill"     

poster <- lego_movie %>%
  html_nodes("#img_primary img") %>%
  html_attr("src")
poster
```


Assignment: Find a web page from a different site and use `rvest` package functions to pull at least one piece of information from that web page.

```{r pirated_movies}
# top 10 most pirated movies of all time; TorrentFreak, Oct 12, 2011
tf <- html("https://torrentfreak.com/top-10-most-pirated-movies-of-all-time-111012/")

# extract with rvest and CSS selectors
titles <- tf %>%
  html_nodes("td:nth-child(2)") %>%
  html_text()
titles

downloads <- tf %>%
  html_nodes("td:nth-child(3)") %>%
  html_text()
downloads

# create a df
pirated <- data.frame(titles, downloads, stringsAsFactors = FALSE)
```


**Top 10 Most Pirated Movies of All Time:**

```{r pirated_table, results = 'asis'}
library(knitr)
kable(pirated, format = "html", row.names = TRUE)
```
*Source: TorrentFreak.com (Oct 2011) via movies downloaded from BitTorrent  (dating back to early 2006)
